The zero--one laws of Kolmogorov and Hewitt--Savage in categorical probability

### 2020-05-03 categorical-zero-one.tex
revised also in light of referee 2

It is known that $\Stoch$ is causal~\cite[Example~11.34]{markov_cats}, and
therefore so are the subcategories $\BorelStoch$ and $\FinStoch$.
\todo[inline]{Added as per reviewer's suggestion, but unsure if it's a good
  inclusion or not.}
\begin{example}
$\CRing_{+}^{\op}$ is \emph{not} causal.
To see this, consider the ring $\mathbb{Z}[\epsilon]/(\epsilon^{2})$.
Elements have the form $a + b\epsilon$, with $a,b\in\bZ$, and multiplication is
defined by the convention $\epsilon^{2} = 0$.
Consider the additive maps given by
\begin{equation}
  f(a + b\epsilon) = & b\\
  g(a + b \epsilon) = & b\\
  h_{1}(a+ b \epsilon) = & a + b \epsilon\\
  h_{2}(a + b \epsilon) = & 0
\end{equation}
Since $gf = 0$, it's clear that the hypothesis holds, so if $\CRing_{+}^{\op}$
were causal, we would have the equation
$g(f(h_{1}(x)y)z) = g(f(h_{2}(x)y)z)$ for all
$x,y,z \in \bZ[\epsilon]/(\epsilon^{2})$.
However, this clearly fails if $z = \epsilon$, $y=1$ and $x = \epsilon$.
\end{example}

@@

\begin{definition}
	\label{defnasequal}
	Given morphisms $p : A \to X$ and $f,g : X \to Y$ in a Markov category, we
    say that $f$ and $g$ are \emph{$p$-a.s.~equal} (or $p$-\emph{almost surely}.~equal) if
	\[
		\tikzfig{asequal}
	\]
	and we also write this more concisely as $f =_{p\as} g$.
  \end{definition}

  This term comes from measure theory, and should be interpreted as meaning
  ``for all $a\in A$, if $x \in X$ is sampled according to $p(a)$, the
  probability that $f(x) \neq g(x)$ is zero''.

@@

	Now let $J = J_1 \,\sqcup\, J_2$ be a decomposition of $J$ into two disjoint subsets having the same cardinality as $J$; this is the step which relies on $J$ being infinite. Let $\tau_1, \tau_2: J \to J$ be injections with images $J_1$ and $J_2$, respectively. Then we claim that
	\[
		\tikzfig{hewsav2}
	\]
	By \Cref{two_infproducts}, it is enough to prove this upon postcomposing with two finite marginalizations $\pi_{F_1}$ and $\pi_{F_2}$. But then using the fact that $\tau_1$ and $\tau_2$ have disjoint image, this follows from the independence assumption, which has
	\[
		\tikzfig{hewsav3}
	\]
	as a special case.

@@

We note that the standard Kolmogorov zero--one law holds for any measurable
spaces, not just for standard Borel spaces - hence this is a strictly less
general statement. As discussed in \cref{infprods_stoch}, it seems unlikely that
the general statement can be proven by applying our machinery to the category
$\Stoch$ of measurable spaces and Markov kernels. This leads to the following problem:

@@

With $\Top$ the category of topological spaces and continuous maps, we consider the \emph{lower Vietoris monad} or \emph{hyperspace monad} $H$ on $\Top$, introduced by Schalk for $T_0$ spaces~\cite{schalk} and investigated generally in~\cite[Section~2]{hyperspace}. Let $\Kl(H)$ be its Kleisli category. Using the symmetric monoidal structure of $H$ guaranteed by~\cite[Corollary~2.53]{hyperspace}, the general construction of Markov categories from symmetric monoidal monads of \cite[Proposition~3.1]{markov_cats} turns $\Kl(H)$ into a Markov category. Concretely, this category has the following structure:
    \begin{enumerate}
        \item The objects are topological spaces.
	\item A morphism $f: X \to Y$ is a continuous function $X \to HY$, where $HY$ denotes the set of closed subsets, equipped with the topology generated by subbasic opens of the form $\{V \subseteq Y \text{ closed} \mid V \cap U \neq \emptyset\}$, where $U \subseteq Y$ ranges over all opens.
	\item Composition is defined as in $\SetMulti$ via~\eqref{relational_composition}, but taking the closure of the union in addition.
        \item The monoidal product is the Cartesian product of topological spaces, and on morphisms
		\[
			(f \tensor g)(x_1,x_2) := f(x_1) \times g(x_2),
		\]
		exactly as in $\SetMulti$.
        \item The comultiplication is defined by $\cop{X}(x) = \cl(\{(x,x)\})$
    \end{enumerate}
\begin{proposition}
    $\Kl(H)$ has Kolmogorov products of any cardinality, given by the usual infinite product of topological spaces with the product topology.
\end{proposition}
\begin{proof}
    First we observe that $H(\prod_{j\in J} X_j)$ is the cofiltered limit of the spaces
    $H(\prod_{j\in F} X_j)$ for $F \subseteq J$ finite.
    To see this, note that the marginalizations induce a continuous map
    	\[
		\alpha \: : \: H\!\left(\prod_{i\in J} X_i\right) \longrightarrow \lim_{F \subseteq J} \, H\!\left(\prod_{i \in J} X_i\right)
	\]
	We show that this is an injection. Thus we start with distinct closed subsets $C, D \subseteq \prod X_i$. We assume without loss of generality that $C \not\subseteq D$, and choose $x \in C \setminus D$.
    Then let $U \ni x$ be an open neighborhood of $x$ not intersecting $D$.
    Since $U$ is open in the product topology, we can assume that there is some finite subset $F \subset J$ so that 
    	\[
		y = (y_i) \in U \quad\Longleftrightarrow\quad y_i \in \pi_i(U) \:\;\forall i\in F.
	\]
    Then $\pi_F(x) \in \pi_F(U)$, which is disjoint from $\pi_F(D)$. Hence $\alpha(C) \neq \alpha(D)$.

@@

	Finally, we must verify that $\alpha$ is an open map.
	It suffices to show that it carries a subbasic open set $\mathrm{Hit}(U) = \{C \subseteq \prod X_i \text{ closed} \mid C \cap U \neq \emptyset\}$ to an open subset of the limit. For a given point $C \in \mathrm{Hit}(U)$, it is enough to find an open neighborhood contained in $\mathrm{Hit}(U)$. But then by making $U$ smaller if necessary, we can assume without loss of generality that $U$ is a basic open in the product topology, meaning that there is a finite $F$ so that
	\[
		y = (y_i) \in U \quad\Longleftrightarrow\quad y_i \in \pi_i(U) \:\; \forall i \in F.
	\]
    Then
	\[
		\alpha(A_U) = \pi_F\inv\left(\left\{V \subseteq \prod_{j \in F} X_F \text{ closed} \:\Bigg|\: V \cap \pi_F(U) \neq \emptyset\right\}\right),
	\]
    which is indeed open in the limit.

>>>>

It is known that $\Stoch$ is causal~\cite[Example~11.34]{markov_cats}, and
therefore so are the subcategories $\BorelStoch$ and $\FinStoch$.

\begin{example}
$\CRing_{+}^{\op}$ is \emph{not} causal.
To see this, consider the ring $\mathbb{Z}[\epsilon]/(\epsilon^{2})$.
Elements have the form $a + b\epsilon$, with $a,b\in\bZ$, and multiplication is
defined by the convention $\epsilon^{2} = 0$.
Consider the additive maps given by
\begin{align*}
	f(a + b\epsilon) = {}& b\\
	g(a + b \epsilon) = {}& b\\
	h_{1}(a+ b \epsilon) = {}& a + b \epsilon\\
	h_{2}(a + b \epsilon) = {}& 0
\end{align*}
Since $gf = 0$, it's clear that the hypothesis holds, so if $\CRing_{+}^{\op}$
were causal, we would have the equation
$g(f(h_{1}(x)y)z) = g(f(h_{2}(x)y)z)$ for all
$x,y,z \in \bZ[\epsilon]/(\epsilon^{2})$.
However, this clearly fails if $z = \epsilon$, $y=1$ and $x = \epsilon$.
\todo[inline]{It would be nice to have such an example. But this one doesn't work, since these additive maps are not unit-preserving.}
\end{example}

@@

\begin{definition}
	\label{defnasequal}
	Given morphisms $p : A \to X$ and $f,g : X \to Y$ in a Markov category, we
	say that $f$ and $g$ are \emph{$p$-a.s.~equal} (or $p$-\emph{almost surely equal}) if
	\[
		\tikzfig{asequal}
	\]
	and we also write this more concisely as $f =_{p\as} g$.
  \end{definition}

  This term comes from measure theory, and should be interpreted as meaning
  that for all those values of $X$ which can occur as outputs of $p$, feeding these values to $f$ and $g$ as inputs is guaranteed to result in the same distribution; at the same time, the behavior of $f$ and $g$ may be different on those input values which do not occur as outputs of $p$.
  \todo[inline]{I've rephrased this in order to make sure that we don't create the impression of making a technically precise statement}

@@

	Now let $J = J_1 \,\sqcup\, J_2$ be a decomposition of $J$ into two disjoint subsets having the same cardinality as $J$; this is the step which relies on $J$ being infinite. Let $\tau_1, \tau_2: J \to J$ be injections with images $J_1$ and $J_2$, respectively. Then we claim that
	\begin{equation}
		\label{hewsav2}
		\tikzfig{hewsav2}
	\end{equation}
	By \Cref{two_infproducts}, it is enough to prove this upon postcomposing with two finite marginalizations $\pi_{F_1}$ and $\pi_{F_2}$. But then using the fact that $\tau_1$ and $\tau_2$ have disjoint image, this follows from the independence assumption, which has
	\[
		\tikzfig{hewsav3}
	\]
	as a special case.

@@

We note that the standard Kolmogorov zero--one law holds for any measurable
spaces, not just for standard Borel spaces, and hence this is a strictly less
general statement. As discussed in \cref{infprods_stoch}, it seems unlikely that
the general statement can be proven by applying our machinery to the category
$\Stoch$ of measurable spaces and Markov kernels. This leads to the following problem:

@@

With $\Top$ the category of topological spaces and continuous maps, we consider the \emph{lower Vietoris monad} or \emph{hyperspace monad} $H$ on $\Top$, introduced by Schalk for $T_0$ spaces~\cite{schalk} and investigated generally in~\cite[Section~2]{hyperspace}. Let $\Kl(H)$ be its Kleisli category. Using the symmetric monoidal structure of $H$ guaranteed by~\cite[Corollary~2.53]{hyperspace}, the general construction of Markov categories from symmetric monoidal monads of \cite[Proposition~3.1]{markov_cats} turns $\Kl(H)$ into a Markov category. Concretely, this category has the following structure:
    \begin{enumerate}
        \item The objects are topological spaces.
	\item A morphism $f: X \to Y$ is a continuous function $X \to HY$, where $HY$ denotes the set of closed subsets, equipped with the topology generated by subbasic opens of the form $\{V \subseteq Y \text{ closed} \mid V \cap U \neq \emptyset\}$, where $U \subseteq Y$ ranges over all opens.
	\item Composition is defined as in $\SetMulti$ via~\eqref{relational_composition}, but taking the closure of the union in addition.
        \item The monoidal product is the Cartesian product of topological spaces, and on morphisms
		\[
			(f \tensor g)(x_1,x_2) := f(x_1) \times g(x_2),
		\]
		exactly as in $\SetMulti$.
        \item The comultiplication is defined by $\cop{X}(x) = \cl(\{(x,x)\})$
    \end{enumerate}
\begin{proposition}
    $\Kl(H)$ has Kolmogorov products of any cardinality, given by the usual infinite product of topological spaces with the product topology.
\end{proposition}
\begin{proof}
    First we observe that $H(\prod_{j\in J} X_j)$ is the cofiltered limit of the spaces
    $H(\prod_{j\in F} X_j)$ for $F \subseteq J$ finite.
    To see this, note that the marginalizations induce a continuous map
    	\[
		\alpha \: : \: H\!\left(\prod_{i\in J} X_i\right) \longrightarrow \lim_{F \subseteq J} \, H\!\left(\prod_{i \in J} X_i\right)
	\]
	We prove that this is a homeomorphism, showing first that it is an injection. Thus we start with distinct closed subsets $C, D \subseteq \prod X_i$. We assume without loss of generality that $C \not\subseteq D$, and choose $x \in C \setminus D$.
    Then let $U \ni x$ be an open neighborhood of $x$ not intersecting $D$.
    By making $U$ smaller if necessary, we can assume without loss of generality that $U$ is a basic open in the product topology, meaning that there is a finite $F$ so that
    \begin{equation}
	    \label{basicU}
		y = (y_i) \in U \quad\Longleftrightarrow\quad y_i \in \pi_i(U) \:\; \forall i \in F.
    \end{equation}
    Then $\pi_F(x) \in \pi_F(U)$, which is disjoint from $\pi_F(D)$. Hence $\alpha(C) \neq \alpha(D)$.

@@

	Finally, we must verify that $\alpha$ is an open map.
	It suffices to show that it carries a subbasic open set $\mathrm{Hit}(U) = \{C \subseteq \prod X_i \text{ closed} \mid C \cap U \neq \emptyset\}$ to an open subset of the limit. For a given point $C \in \mathrm{Hit}(U)$, it is enough to find an open neighborhood contained in $\mathrm{Hit}(U)$. 
	But then again by choosing smaller $U$ if necessary, so that $C \cap U \neq \emptyset$ still holds, we can assume~\eqref{basicU} without loss of generality.
    Then
	\[
		\alpha(A_U) = \pi_F\inv\left(\left\{V \subseteq \prod_{j \in F} X_F \text{ closed} \:\Bigg|\: V \cap \pi_F(U) \neq \emptyset\right\}\right),
	\]
    which is indeed open in the limit.
